--- 
title: "Analytics Portfolio"
author: "George Woolsey"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: 
  bookdown::gitbook: default
documentclass: book
bibliography: [packages.bib]
biblio-style: apalike
# link-citations: yes
description: "Using the bookdown package to write a book of analytics projects. The output format for this example is bookdown::gitbook."
---

# Prerequisites

This is a _sample_ book written in **Markdown**. You can use anything that Pandoc's Markdown supports, e.g., a math equation $a^2 + b^2 = c^2$.

The **bookdown** package can be installed from CRAN or Github:

```{r eval=FALSE}
install.packages("bookdown")
# or the development version
# devtools::install_github("rstudio/bookdown")
```

Remember each Rmd file contains one and only one chapter, and a chapter is defined by the first-level heading `#`.

To compile this example to PDF, you need XeLaTeX. You are recommended to install TinyTeX (which includes XeLaTeX): <https://yihui.name/tinytex/>.

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(),
  'bookdown'
  , 'knitr'
  , 'rmarkdown'
  , 'car'
  , 'dataRetrieval'
  , 'dygraphs'
  , 'ggthemes'
  , 'gplots'
  , 'LAGOSNE'
  , 'leafpop'
  , 'lubridate'
  , 'mapview'
  , 'pdftools'
  , 'R.matlab'
  , 'RApiSerialize'
  , 'RColorBrewer'
  , 'readxl'
  , 'rnassqs'
  , 'rvest'
  , 'scales'
  , 'sf'
  , 'tidyr'
  , 'tidyverse'
  , 'USAboundaries'
  , 'USAboundariesData'
  , 'viridis'
  , 'xgboost'
  , 'xts'
), 'packages.bib')
```
