--- 
title: "Analytics Portfolio"
author: "George Woolsey"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: 
  bookdown::gitbook: default
documentclass: book
bibliography: [packages.bib]
biblio-style: apalike
# link-citations: yes
description: "Using the bookdown package to write a book of analytics projects. The output format for this example is bookdown::gitbook."
---

# Introduction

My name is George Woolsey and I am currently pursuing on a MS in Forest Sciences at Colorado State University. I am working with Dr. Wade Tinkham in the [Forest Biometrics Lab](https://sites.warnercnr.colostate.edu/wtinkham/). Our lab utilizes a combination of field inventories, remote sensing, GIS, and forest growth/landscape change models to characterize and address a wide array of applied forest management questions.

## This book
This book of selected analytics projects was compiled using **R Markdown** and the **bookdown** package which can be installed from CRAN or Github (code block below). The majority of these projects were completed as part of the Environmental Data Science course in the spring of 2022.

```{r eval=FALSE}
install.packages("bookdown")
# or the development version
# devtools::install_github("rstudio/bookdown")
```


```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(),
  'bookdown'
  , 'knitr'
  , 'rmarkdown'
  , 'car'
  , 'dataRetrieval'
  , 'dygraphs'
  , 'ggthemes'
  , 'gplots'
  , 'LAGOSNE'
  , 'leafpop'
  , 'lubridate'
  , 'mapview'
  , 'pdftools'
  , 'R.matlab'
  , 'RApiSerialize'
  , 'RColorBrewer'
  , 'readxl'
  , 'rnassqs'
  , 'rvest'
  , 'scales'
  , 'sf'
  , 'tidyr'
  , 'tidyverse'
  , 'USAboundaries'
  , 'USAboundariesData'
  , 'viridis'
  , 'xgboost'
  , 'xts'
), 'packages.bib')
```
